---
title: "Getting start"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(dads)
set.seed(4242)
```

# Getting started

## The simplest of all analysis: simulating diversity only

One of the simplest things to do with the `dads` package is just to simulate a birth death tree.
For that you can use the function `dads` and specify your stopping rule.
The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:

 * `"max.taxa"   = n` stop when `n` taxa are generated;
 * `"max.living" = n` stop when there is `n` co-occuring taxa of the same age (i.e. "living" taxa);
 * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);

For example, we might want to generate a birth-death tree with 20 taxa:

```{r}
## Setting a stopping rule to reach a maximum of 20 taxa
my_stop_rule <- list(max.taxa = 20)
```

We can now run the simulations using:

```{r}
## Running the birth death simulation
my_tree <- dads(stop.rule = my_stop_rule)
```

> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`. The simulation will then stop when either of these conditions are met.

The resulting object is a classic `"phylo"` object that you can simply plot or visualise like so:

```{r}
## The tree object
my_tree
## Plotting it
plot(my_tree)
```

### Changing the birth-death parameters

People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
By default `dads` runs a pure birth model (the speciation is set to 1 and the extinction to 0).
However, you can easily change that by specifying your new birth death parameters:

```{r}
## my birth death parameters
my_params <- list(speciation = 1,
                  extinction = 1/3)
```
> Note that here it is not necessary to specify `extinction = 1` since this is the default option, you can always just change the parameter of interest (e.g. changing `extinciton = 0` to `extinction = 1/3`). However, we think it's good practice to attribute both parameters specifically to avoid any confusion.

You can then run the same birth death tree with extinction:

```{r}
## Generating a birth death tree with extinctions:
my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
## Visualising the new tree
plot(my_tree)
```

## Slightly more complex: simulating disparity and diversity

Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth death models).
Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

`"traits"` objects contain one or more processes which are the ways to generate the trait.
The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion).
This is used by default with the `make.traits` function:

```{r}
## Creating the traits object
my_trait <- make.traits()
```

This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):

```{r}
## Which process is in here?
my_trait
## What does it look like?
plot(my_trait)
```

By default, this trait is called "A".
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `dads` function to generate the tree and the traits:

```{r}
## Simulate disparity and diversity
my_data <- dads(bd.params = my_params,
                stop.rule = my_stop_rule,
                traits    = my_trait)
```

Et voilà! We now have a simple disparity and diversity simulation.
We can see what's in the results by simply printing it or plotting it:

```{r}
## What's in there
my_data
## Plotting the disparity and diversity
plot(my_data)
```

You can then extract the components you need for your specific analysis like so:

```{r}
## Extracting the tree (a "phylo" object)
the_generated_tree <- my_data$tree
class(the_generated_tree)

## Extracting the data (a "matrix")
the_generated_data <- my_data$data
class(the_generated_data)
```

## Slightly more more complex: simulating linked disparity and diversity 

The example above is also still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e. the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
It is possible to add this aspect using `"modifiers"` objects.
`"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.

