---
title: "Getting start"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# A simple analyses


## What you need?


 * The speciation and extinction parameter

 ```{r, eval = FALSE}
## The birth death parameters
my_params <- list(speciation = 1,
                  extinction = 1/3)
```

 * A stopping rule

 ```{r, eval = FALSE}
## Creating a stopping rule
my_stop_rule <- list(max.taxa = 15)
```

 * A "traits" object

 ```{r, eval = FALSE}
## A simple Brownian motion object
my_bm_trait <- make.traits()
```


## How to simulate disparity (traits) and diversity (taxa)?

 ```{r, eval = FALSE}
## Running a pretty basic simulation
simple_dads <- dads(bd.params = my_params,
                    stop.rule = my_stop_rule,
                    traits    = my_bm_traits)
```



> Note that you can also simply simulate diversity by setting the traits argument to `NULL`. This will simply simulate a birth death tree. If this is your objective, you'd probably be better off using one of the excellent `R` packages dedicated to this task like `TreeSim`.