---
title: "`make.traits`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Making trait processes with `make.traits()` {#maketraits}

```{r, echo = FALSE}
set.seed(123)
```

## The process (`process`)

The function `make.traits` allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
Note that it _can_ depend on both the ancestor and the branch length but does _not necessary needs_ (i.e. the process can be only based on the previous state or only on branch length or on neither).

Trait processes in `dads` are functions that must always intake the following arguments by default.

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length value
 * `...`: a placeholder for any extra arguments

For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):

```{r}
## A valid (but useless) process
valid.process <- function(x0, egde.length = 1, ...) {
    return(42)
}
```

> Note that the argument `edge.length` is set to `1` by default. In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speeding up the `dads` function).

On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):

```{r}
## A wrongly formated process
invalid.process <- function(x0) {
    return(rnorm(1, mean = x0))
}
```

The `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.

Once a process is chosen, you can feed it to the `make.traits` function:

```{r}
## Creating a trait object
my_trait_object <- make.traits(process = BM.process)
```

This creates `"dads"` `"traits"` objects that you can print, and visualise using the `plot` function:

```{r}
## The class of the object
class(my_trait_object)
## What's in it?
my_trait_object
## What does the process looks like
plot(my_trait_object)
```

Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using:


```{r}
## What's actually in that object?
print.dads(my_trait_object, all = TRUE)
```

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.

Note that it is possible to make `"traits"` objects with multiple processes (that can be the same):

```{r}
## 4 traits: two BM, one OU and one normal non process
four_traits <- make.traits(process = c(BM.process,
                                       BM.process,
                                       OU.process,
                                       no.process))
four_traits
```

You can visualise them individually using the `trait` argument in `plot.dads`:

```{r}
## Plot options (4 plots in one window)
par(mfrow = c(2,2))
plot(four_traits, trait = 1)
plot(four_traits, trait = 2)
plot(four_traits, trait = 3)
plot(four_traits, trait = 4)
```

## The number of traits `n` and the starting values `start`

Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
By default they are set to `n = 1` and `start = 0`.
This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
It is however possible to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value `1`:

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = 1)
```

Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = c(1,2,3))
```

Note that the number of traits are distributed per processes.
If the traits contains multiple process, the number of traits are distributed per processes:

```{r}
## two 3D processes (BM and OU)
make.traits(c(BM.process, OU.process), n = 3)
## one 1D processes (BM) and one 4D process (OU)
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

And starting values are distributed for all the traits or for the traits one by one:

```{r}
## two 3D processes (BM and OU) starting with 1
make.traits(c(BM.process, OU.process), n = 3, start = 1)
## two 3D processes (BM and OU) starting with values 1 to 6
make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
## two 3D processes (BM and OU) with the two first ones starting
## with 1 and the 4 other ones with the default (0)
make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
```

## Extra argument for the processes with `process.args`

You can also feed extra arguments to your process(es) functions. For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can intake a specific random number generator as a function:

```{r}
## no process trait using the normal distribution (default)
make.traits(no.process, process.args = list(fun = rnorm))
## no process trait using the uniform distribution
## bounded between 1 and 100
make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
```

You can also add multiple extra arguments for multiple processes giving them as a list.

```{r}
## Two traits with no process:one normal and one uniform (1,100)
make.traits(process      = c(no.process, no.process),
            process.args = list(list(fun = rnorm),
                                list(fun = runif, min = 1, max = 100)))
```

If one process do not need extra argument you must still give it and extra `NULL` process argument:

```{r}
## Three traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)))
```

## Naming the traits with `trait.names`

As traits become more and more complex, it can be useful to give clearer names to each process.
This is easily done using the `trait.names` argument that attributes one name per process:

```{r}
## A simple trait with a proper name
simple_trait <- make.traits(trait.names = "1D Brownian Motion")
simple_trait
```

This becomes more useful if we use the complex example above:

```{r}
## Three named traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)),
            ## Naming each trait
            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
```


## Combining multiple traits with `add`

You can also add traits to already existing trait objects using the simple `add` option.
This option just intakes a `"dads"` `"traits"` object and the additional process(es) will be added to it. For example:

```{r}
## Creating on simple default Brownian motion
one_process <- make.traits(trait.names = "BM")

## Creating a new trait (a 3D OU.process)
## and adding the previous one
two_processes <- make.traits(OU.process, n = 3, add = one_process,
                             trait.names = "3D OU")

## Only one process
one_process

## The two processes
two_processes
```


## Testing the traits with `test`

This bit is more for development.
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.


## Templates for making your very own process

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `dads` processes to set all the arguments but `x0` with default values (just in case).
Also, note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but are rather generalised/simplified version that you can use as a template

### A simple Brownian Motion process template

```{r}
## A simple Brownian motion process
my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
    ## Drawing a random number from a normal distribution
    ## with x0 as the and a given standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = x0, sd = sd * edge.length)

    ## Return the number
    return(result)
}
```

### A simple Ornstein-Uhlenbeck process template

```{r}
## A simple Ornstein-Uhlenbeck motion process
my.OU.process <- function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
    ## Calculate the mean based on alpha
    mean <- x0 * exp(-alpha)
    ## Calculate the standard deviation based on alpha and the variance
    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
    ## Draw a random number from a normal distribution
    ## using this mean and standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = mean, sd = sd * branch.length)

    ## Return the number
    return(result)
}
```



