---
title: "`make.traits`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../references.bib]
output:
  html_document: default
  pdf_document: default
---

# Simulating traits {#maketraits}

```{r, echo = FALSE}
library(treats)
set.seed(123)
```

In `treats`, traits are simulating by providing a `traits` object to the `traits` argument.
This object is generated using `make.traits`.

> **What is a trait in biology?** This is one of these epistemological questions like what _is_ a species and is not covered in this package.
For the purpose of _this manual_, a **biological trait** can be any coherent character of any number of dimensions which may or may not be independent of other traits.
Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
For the purpose of the syntax in _this package_ a **trait in `treats`** is a column in the output trait matrix.
For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
This single "beak shape" **biological trait** is then composed of 912 **traits in `treats`** (i.e. a matrix of $(4+300)\times3$ columns).

## The process (`process`)

The function `make.traits` allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules needed to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
Note that it _can_ depend on both the ancestor and the branch length but does _not necessarily need to_, i.e. the process can be based only on the previous state or only on branch length or on neither.

### The syntax (how to code a process?)

Trait processes in `treats` are functions that must always take the following arguments by default.

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length
 * `...`: a placeholder for any extra arguments

For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
In this example, the process is not dependent on either the previous state (`x0`) or the branch length (`edge.length`).

```{r}
## A valid (but useless?) process
valid.process <- function(x0 = 0, edge.length = 1, ...) {
    return(42)
}
```

> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `treats` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):

```{r}
## A wrongly formatted process
invalid.process <- function(x0 = 0) {
    return(rnorm(1, mean = x0))
}
```

This will not work in `make.traits` (see below).

### Using a `"process"` in `treats`

You can design your own process as a function, as long as it has a valid syntax.
Alternatively, the `treats` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.

Once a process is chosen, you can feed it into the `make.traits` function:

```{r}
## Creating a trait object
my_trait_object <- make.traits(process = BM.process)
```

This creates `"treats"` `"traits"` objects that you can print and visualise using the `plot` function:

```{r}
## The class of the object
class(my_trait_object)
## What's in it?
my_trait_object
## What does the process looks like
plot(my_trait_object)
```

Note that you can see the multiple options for plotting the trait process by looking at `?plot.treats` manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `treats` package):


```{r}
## What's actually in that object?
print.treats(my_trait_object, all = TRUE)
```

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the `"traits"` object.

Note that it is possible to make `"traits"` objects with multiple traits and multiple processes (that can be the same or different for each trait):

```{r}
## Four traits: two BM, one OU and one normal non process
four_traits <- make.traits(process = c(BM.process,
                                       BM.process,
                                       OU.process,
                                       no.process))
four_traits
```

You can visualise them individually using the `trait` argument in `plot.treats`:

```{r}
## Plot options (4 plots in one window)
par(mfrow = c(2,2))
plot(four_traits, trait = 1)
plot(four_traits, trait = 2)
plot(four_traits, trait = 3)
plot(four_traits, trait = 4)
par(mfrow = c(1,1))
```

## The number of traits `n` and the starting values `start`

Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
By default they are set to `n = 1` and `start = 0`.
This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
It is possible however, to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimension starting with the value `1`:

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = 1)
```

Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = c(1,2,3))
```

Note that the number of traits are distributed per processes.
If the traits contains multiple processes, the number of traits are distributed per processes:

```{r}
## two 3D processes (BM and OU)
make.traits(c(BM.process, OU.process), n = 3)
## one 1D processes (BM) and one 4D process (OU)
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

And starting values are distributed for all the traits or for the traits one by one:

```{r}
## two 3D processes (BM and OU) starting with 1
make.traits(c(BM.process, OU.process), n = 3, start = 1)
## two 3D processes (BM and OU) starting with values 1 to 6
make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
## two 3D processes (BM and OU) with the two first ones starting
## with 1 and the 4 other ones with the default (0)
make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
```

### What is a trait in `treats`?

Because it would be impossible to accommodate all definitions of a trait in `treats` we chose an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
In `treats`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `treats`:

```{r, eval = FALSE}
## Three traits with one process:
make.traits(BM.process, n = 3, start = c(1,2,3))
## Six traits with two processes:
make.traits(c(BM.process, OU.process), n = 3)
## Five traits with two processes
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

## Extra argument for the processes with `process.args`

You can also feed extra arguments to your process(es) functions. 
For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can take a specific random number generator as a function:

```{r}
## no process trait using the normal distribution (default)
make.traits(no.process, process.args = list(fun = rnorm))
## no process trait using the uniform distribution
## bounded between 1 and 100
make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
```

You can also add multiple extra arguments for multiple processes giving them as a list.

```{r}
## Two traits with no process:one normal and one uniform (1,100)
make.traits(process = c(no.process, no.process),
            process.args = list(list(fun = rnorm),
                                list(fun = runif, min = 1, max = 100)))
```

If one process does not need extra argument you must still give it an extra `NULL` process argument:

```{r}
## Three traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)))
```

## Naming the traits with `trait.names`

As traits become more and more complex, it can be useful to give clearer names to each process.
This is easily done using the `trait.names` argument that attributes one name per process:

```{r}
## A simple trait with a proper name
simple_trait <- make.traits(trait.names = "1D Brownian Motion")
simple_trait
```

This becomes more useful if we use the complex example above:

```{r}
## Three named traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)),
            ## Naming each trait
            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
```


## Combining multiple traits with `add`

You can also add traits to already existing trait objects using the simple `add` option.
This option just takes a `"treats"` `"traits"` object and the additional process(es) will be added to it. 
For example:

```{r}
## Creating a simple default Brownian motion
one_process <- make.traits(trait.names = "BM")

## Creating a new trait (a 3D OU.process)
## and adding the previous one
two_processes <- make.traits(OU.process, n = 3, add = one_process,
                             trait.names = "3D OU")

## Only one process
one_process

## The two processes
two_processes
```

## Using a background trait

`traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
This allows traits to be simulated for _all_ tips whenever a trait is generated for one tip.
This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
The `background` argument takes any output from the `make.traits` function in a nested way:

```{r}
## Generating a default BM trait:
BM_trait <- make.traits()
## Generating an OU trait with a background BM trait
my_trait <- make.traits(process = OU.process, background = BM_trait) 
```

Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
However, you should always make sure that the background trait has the same dimensions as the main trait.
When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
You can remove these nodes using the [`drop.things` function](#dropthings).

```{r}
set.seed(1)
## Generating a pure birth tree with the background trait
tree_bkg <- treats(stop.rule = list(max.taxa = 20),
                 traits = my_trait)
## This tree has many internal singleton nodes
plot(tree_bkg)
```

## Saving trait values at different time steps

You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `treats`.
This will apply the `traits` object to all lineages currently alive at the required time steps.
These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.

```{r}
set.seed(123)
## Generating a birth-death tree with a BM trait and saving steps at specific times
tree_steps <- treats(stop.rule  = list(max.time = 3),
                   bd.params  = list(speciation = 1, extinction = 0.1),
                   traits     = make.traits(),
                   save.steps = c(1/3, 1, 2))
## This also creates internal singleton nodes
plot(tree_steps)
abline(v = 3 - c(1/3, 1, 2), lwd = 0.5, col = "grey")
```

## Traits implemented in `treats`

If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `treats`. You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:

 * `BM.process`: this is the well known [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).

```{r, echo = FALSE}
plot(make.traits(process = BM.process), main = "The Brownian motion")
```

 * `OU.process`: this is the equally famous [Ornstein–Uhlenbeck process](https://en.wikipedia.org/wiki/Ornstein%E2%80%93Uhlenbeck_process).

```{r, echo = FALSE}
plot(make.traits(process = OU.process), main = "The Ornstein–Uhlenbeck process")
```

 * `no.process`: this process has... no process. In other words, this is a non time dependent process; the simulated value does not depends on the ancestors' value nor the branch length. It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc.


```{r, echo = FALSE}
plot(make.traits(process = no.process, process.args = list(fun = runif)), main = "No process using runif\n(yes, it \"looks\" like an OU!)")
```

 * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local long-term mean values. The default OU process has one long-term mean towards which the values are drawn with the alpha parameter (the elastic band). The single long-term mean is usually 0. However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.

```{r, echo = FALSE}
plot(make.traits(multi.peak.process, process.args = list(peaks = c(1, 5, 10))), main = "A multi-peak (OU) process\nwith peaks (optima) at 1, 5 and 10")
```

 * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating gradually through time, new trait values are more likely to be different to the traits of their ancestor following a `repulsion` parameter.

```{r, echo = FALSE}
plot(make.traits(repulsion.process, process.args = list(repulsion = 5)), main = "A repulsion process\nwith a repulsion parameter of 5")
```

## Testing the traits with `test`

**This bit is still in development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `treats`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.


## Templates for making your very own process

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `treats` processes to set all the arguments with default values (just in case).

> Note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but an easier to edit version that you can use as a template:

### A simple Brownian Motion process template

```{r}
## A simple Brownian motion process
my.BM.process <- function(x0 = 0, edge.length = 1, sd = 1, ...) {
    ## Drawing a random number from a normal distribution
    ## with x0 as the and a given standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = x0, sd = sqrt(sd^2 * edge.length))

    ## Return the number
    return(result)
}
```

### A simple Ornstein-Uhlenbeck process template

```{r}
## A simple Ornstein-Uhlenbeck motion process
my.OU.process <- function(x0 = 0, edge.length = 1, var = 1, alpha = 1, ...) {
    ## Calculate the mean based on alpha
    mean <- x0 * exp(-alpha)
    ## Calculate the standard deviation based on alpha and the variance
    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
    ## Draw a random number from a normal distribution
    ## using this mean and standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = mean, sd = sqrt(sd^2 * edge.length))

    ## Return the number
    return(result)
}
```



