---
title: "`plot.dads`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
# bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(dads)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
options(rgl.useNULL = TRUE) # Suppress the separate window.
```

# Plots


`"dads"` objects can be directly plotted in `dads` using the S3 `plot.dads` function (or just `plot(x)` if `x` is of class `"dads"`).


## Plotting traits

`"dads"` `"traits"` objects are covered in the [traits section](#maketraits).
You can use `plot.dads` to plot them by choosing which specific trait to plot using the `trait` argument (default is 1):

```{r, fig.height=8, fig.width=8}
## Making a list of three traits
list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
## Plotting each trait separately
par(mfrow = c(2, 2))
plot(list_of_traits, trait = 1)
## Using different colours options
plot(list_of_traits, trait = 2, col = c("red", "purple", "pink"))
## Not using the default plot name
plot(list_of_traits, trait = 3, main = "OU process",
     cent.tend = sd, quantiles = c(10, 30))
```

You can also control the number of replicates in the simulation by using the `simulations` option (the default is 50).
Bigger numbers leads to more time but smoother looking plots while smaller ones are more stochastic:

```{r, fig.width=6, fig.height=12}
par(mfrow = c(2,1))
plot(list_of_traits, trait = 2, simulations = 10, main = "10 BM simulations")
plot(list_of_traits, trait = 2, simulations = 1000, main = "1k BM simulations")
```


## Plotting dads results

If `dads` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
You'll find all the tree plotting option in the `?plot.phylo` manual page.

```{r}
## A simple pure birth tree
my_tree <- dads(stop.rule = list(max.taxa = 20))
plot(my_tree, main = "Plotting a \"phylo\" object")
```

However, if you also simulated a trait along with the tree you can use the `plot.dads` function to plot both the tree and the trait:

```{r}
## A simple pure birth tree with a BM process
my_tree <- dads(stop.rule = list(max.taxa = 20), traits = make.traits())
## Playing with the default options
plot(my_tree, main = "A tree and traits")
```

By default, elements are coloured as follows: nodes and tips are points and coloured in blue if they are tips (light blue if they are fossils) and nodes are in orange.
Branches linking them are grey lines.
You can of course change this colour palette to something of you preference by calling the normal arguments that can be passed to `points` or `lines`. For example `pch` for the point type or `lty` for the line type.
Furthermore, time is plotted conventionally from left to right (left is towards the past, right is towards the present) but you can change that by specifying `xlim`.


```{r, fig.height = 6, fig.width = 6}
## Playing with some more options
plot(my_tree, main = "A tree and traits",
     ## Changing nodes colours and type
     col = c(tips = "pink", nodes = "purple"), pch = 21,
     ## Changing edges colour and type
     lwd = 2, lty = 3, edges = "yellow",
     ## Changing the x axis orientation and label
     xlim = c(0, 2.5), xlab = "Time into the past")
```

> Note that you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g. `col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g. `col = grDevices::heat.colors` will use the specified function to change the elements colours with time).

And you can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).

```{r, fig.height = 6, fig.width = 6}
## Adding the default legend
plot(my_tree, legend = TRUE)
```

### 3D version!

Of course, if you're simulating multiple traits, you can always plot different ones.

```{r, fig.height = 6, fig.width = 6}
## Specifying a 3D trait process
my_3D_trait <- make.traits(n = 3, )
## Simulating a birth death tree with that 3D trait
my_data <- dads(bd.params = list(extinction = 0.2),
                stop.rule = list(max.living = 50),
                traits    = my_3D_trait)
```

You can toggle which trait to plot using the `trait` option.
Either by providing a single value to plot that specific trait against time or by providing two traits.

```{r, fig.height = 12, fig.width = 6}
par(mfrow = c(2,1))
## Plotting the second trait
plot(my_data, trait = 2, main = "Trait 2")
## Plotting the correlation between trait 1 and 2
plot(my_data, trait = c(1,2), main = "Traits 1 and 2")
```

As mentioned above, the `col` argument can take a function for scaling the elements colours with time.
This can be useful for adding time a third dimension to these 2D plots:

```{r}
## Plotting the correlation between trait 1 and 2
## with time as a 3rd dimensions
plot(my_data, trait = c(1,2), main = "Traits 1 and 2",
     col = grDevices::heat.colors, legend = TRUE,
     ## Highlighting the tips in black for visibility
     tips.nodes = "black")
```

But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
This is done through the `rgl` package and activated in `dads` using the option `use.3D = TRUE`.

If two traits are provided, the 3rd dimensions is going to be time by default:

```{r}
## Plotting the tree and 2 traits in 3D: woah!
plot(my_data, trait = c(1, 2), use.3D = TRUE)
rglwidget()
```

If three traits are provide, you can set up a fourth coloured dimension as time (similarly to the 2D plots with the colour gradients).
This uses the `rgl` functions, namely `lines3d`, `points3d` (if `type = "p"`; default) or `sphere3d` (if `type = "s"`).
The additional arguments `...` are directly handled and attributed to the corresponding function.

```{r}
## Plotting the tree and 3 traits in 3D
plot(my_data, trait = c(1, 2, 3), use.3D = TRUE,
     col = grDevices::heat.colors, type = "s", radius = 0.1)
rglwidget()
```
