--- 
title: "dads manual"
author: "Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../references.bib]
link-citations: yes
github-repo: TGuillerme/dads
description: "dads R package vignette"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `dads`

Disparity and diversity simulations.

The core of this package is based on diversitree function blabalba


## What is **not** `dads`s?

Link to RPANDA, diversitree, etc.

## What is `dads`?

Blablaba

### Modular?

Blablaba

## Installing and running the package

You can install this package easily, directly from the CRAN:

```{r, eval = FALSE}
install.packages("dads")
```

Alternatively, for the most up to data version and some functionalities not compatible with the CRAN, you can use the package through GitHub using `devtool` (see [to CRAN or not to CRAN?](#noCRAN) for more details):

```{r, eval = FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

## Installing the latest released version directly from GitHub
install_github("TGuillerme/dads", ref = "release")
```

```{r, eval = FALSE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dads)
## Setting a random seed for repeatability
set.seed(123)
```

## Help

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).

<!-- 
## Citations

To cite the package, this manual or some specific functionalities, you can use the following references:

Furthermore, don't forget to cite `R`:

> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:

> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.
 -->

<!-- ### Why is it important to cite us?

Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?

All the people involve in making the `dads` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
I created the package with this idea in mind and I am still sticking to it.
However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dads` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!

Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!

Thank you! -->
<!-- ## Associated papers -->

<!-- Here are some papers that have used the `dispRity` package: -->
